---
title: "final project proposal"
author: "Waheeb Algabri, Joe Garcia, Lwin Shwe, Mikhail Broomes"
output:
  html_document:
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
  pdf_document: default
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


```{r libraries, message=FALSE, echo=FALSE}
library(tidyverse)
library(modelr)
library(glue)
library(tidyverse)
library(DataExplorer)
library(knitr)
library(cowplot)
library(finalfit)
library(correlationfunnel)
library(ggcorrplot)
library(RColorBrewer)
library(naniar)
library(mice)
library(MASS)
select <- dplyr::select
library(kableExtra)
library(car)
library(glmtoolbox)
library(pROC)
library(caret)
library(robustbase)
library(corrplot)
```


## Introduction:

Our final project aims to analyze the predictive power of Department of Education (DOE) ratings on the academic performance of New York City (NYC) high schools. Leveraging data from [OpenData's - School Quality Report](https://data.cityofnewyork.us/Education/2020-2021-School-Quality-Reports-High-School/26je-vkp6/about_data) for NYC high schools between 2020 and 2021. Our final project will potentially include more up-to-date educational information and geographic data to augment our analysis.

First, let's read in our source CSV file. This is posted in our [GitHub repository](https://github.com/waheeb123/Data-621) in the interest of reproducibility.

```{r data-read-in}
df <- read.csv("https://raw.githubusercontent.com/waheeb123/Data-621/main/Final%20Project/2020-2021_School_Quality_Reports_-_High_School_20240407.csv")
```

There are several predictor variables of interest to us. The NYC School Quality Report from the Department of Education included ratings for schools in the following categories. We would like to see how effective these ratings are at predicting student success:

- *Quality Review Rating*
- *Achievement Rating*
- *Environment Rating*
- *College and Career Readiness Rating*


In addition, it'd be interesting to see if these ratings could be replaced by proxy variables? Doing so could save the Department of Education (DOE) time in not assigning ratings when they could have similar impact by knowing certain values for a school.

Our response variable will be the average student's SAT Score at a given school. SAT Scores are an imperfect metric given their [correlation with other socioeconomic factors](https://files.eric.ed.gov/fulltext/ED563073.pdf), but for our purposes can serve as an imperfect benchmark to measure academic performance.

**Main research question:** Are DOE ratings effective predictors of a school's ability to cultivate high academic achievement in students, and what alternative proxy variables offer improved accuracy in predicting academic performance, as assessed by SAT scores?



## EDA

First, let's plot the counts of different rating combinations by school. This should give us a sense if schools with one rating for a given bucket tend to have similar ratings for other categories (i.e., schools with high *Quality Review Rating* values tend to have high *Achievement Ratings* as well).

|`survey_pp_RI`|Rigorous Instruction - Percent Positive|
|`survey_pp_CT`|Collaborative Teachers - Percent Positive|
|`survey_pp_SE`|Supportive Environment - Percent Positive|
|`survey_pp_ES`|Effective School Leadership - Percent Positive|
|`survey_pp_SF`|Strong Family-Community Ties - Percent Positive|
|`survey_pp_TR`|Trust - Percent Positive|
|`QR_1_1`|Quality Review - How interesting and challenging is the curriculum?|
|`QR_1_2`|Quality Review - How effective is the teaching and learning?|
|`QR_2_2`|Quality Review - How well does the school assess what students are learning?|
|`QR_3_4`|Quality Review - How clearly are high expectations communicated to students and staff?|
|`QR_4_2`|Quality Review - How well do teachers work with each other?|
|`QR_1_4`|Quality Review - How safe and inclusive is the school while supporting social-emotional growth?|
|`QR_1_3`|Quality Review - How well does the school allocate and manage resources?|
|`QR_3_1`|Quality Review - How well does the school identify, track, and meet its goals?|
|`QR_4_1`|Quality Review - How thoughtful is the school’s approach to teacher development and evaluation?|
|`QR_5_1`|Quality Review - How well are school decisions evaluated and adjusted?|
|`n_nocat_cri_4yr_all`|N Count - 4-Year College Readiness Index (without CAT)|
|`val_nocat_cri_4yr_all`|Metric Value - 4-Year College Readiness Index (without CAT)|
|n_mean_score_act_engl_all|N Count - Average Score ACT English|
|val_mean_score_act_engl_all|	Metric Value - Average score of students in the current cohort who took the ACT English exam|
|`n_mean_score_act_math_all`|	N Count - Average Score ACT Math|
|`val_mean_score_act_math_all`|	Metric Value - Average score of students in the current cohort who took the ACT Math exam|
|`n_mean_score_act_read_all`|	N Count - Average Score ACT Reading|
|`val_mean_score_act_read_all`|	Metric Value - Average score of students in the current cohort who took the ACT Reading exam|
|`n_mean_score_act_scie_all`|	N Count - Average Score ACT Science|
|`val_mean_score_act_scie_all`|	Metric Value - Average score of students in the current cohort who took the ACT Science exam|
|`n_mean_score_sat_math_all`|	N Count - Average Score SAT Math|
|`val_mean_score_sat_math_all`|	Metric Value - Average score of students in the current cohort who took the SAT Math exam|
|`n_mean_score_sat_writ_all`|	N Count - Average Score SAT Writing|
|`val_mean_score_sat_writ_all`|	Metric Value - Average score of students in the current cohort who took the SAT Reading and Writing exam|

## Exploratory Data Analysis:

First, let's plot the counts of schools receiving all possible different rating combinations. This should give us a sense whether schools with a rating for one category tend to have a similar rating for other categories. That is, schools with high *Quality Review Rating* values might be expected to have high *Achievement Ratings*, and this plot will highlight whether expectations for these variables match reality for all possible buckets. 


## Modeling






## Further Work
Below are some additional modeling points to consider for the final project, as well as to augment our dataset for our final project:

- SAT Scores are not a perfect indicator of future academic [<Citation here>]. Using other response variables could paint a more complete picture on the educational variables that most impact outcomes
- Include other educational outcome metrics (job placement rates on graduation, income by school) joined to our data
- Fortunately, NYC's DBN (*District-Borough Number*) system allows for easier joining to other education datasets posted on NYC Open Data
- Use a more recent dataset than 2013-2014. NYC Open Data is an excellent tool for this








