---
title: "Homework-3"
author: "Waheeb Algabri, Joe Garcia, Lwin Shwe, Mikhail Broomes"
output:
  html_document:
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
  pdf_document: default
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)

library(tidyverse)
library(modelr)
library(DataExplorer)
library(correlationfunnel)
library(caret)
library(knitr)
library(confintr)
library(psych)
library(car)
library(corrplot)
library(RColorBrewer)
library(MASS)
select <- dplyr::select
library(glmtoolbox)
library(cowplot)
library(pROC)
library(bestglm)

```


## Homework 3 - Logistic Regression

### Analysis of Factors that Influence Crime Rates:

In this homework assignment, we'll delve into examining, dissecting, and modeling a dataset that provides insights into crime across different neighborhoods within a bustling metropolis. Each entry in the dataset includes a variable denoting whether the crime rate exceeds the median value `(1)` or falls below it `(0)`
Our goal is to construct a binary logistic regression model using the training dataset. This model aims to forecast whether a neighborhood is prone to experiencing elevated levels of crime.
Here's a brief rundown of the key variables in the dataset:

|Column|Description|
|--|--|
|`zn`|proportion of residential land zoned for large lots (over 25000 square feet) (*predictor variable*)|
|`indus`|proportion of non-retail business acres per suburb (*predictor variable*)|
|`chas`|a dummy var. for whether the suburb borders the Charles River (1) or not (0) (*predictor variable*)|
|`nox`|nitrogen oxides concentration (parts per 10 million) (*predictor variable*)|
|`rm`|average number of rooms per dwelling (*predictor variable*)|
|`age`|proportion of owner-occupied units built prior to 1940 (*predictor variable*)
|`dis`|weighted mean of distances to five Boston employment centers (*predictor variable*)
|`rad`|index of accessibility to radial highways (*predictor variable*)|
|`tax`|full-value property-tax rate per $10,000 (*predictor variable*)|
|`ptratio`|pupil-teacher ratio by town (*predictor variable*)|
|`lstat`|lower status of the population (percent) (*predictor variable*)|
|`medv`|median value of owner-occupied homes in $1000s (*predictor variable*)|
|**`target`**|**whether the crime rate is above the median crime rate (1) or not (0) (*response variable*)**|

```{r load-train-data, echo=F}
train_df <- read.csv("https://raw.githubusercontent.com/waheeb123/Data-621/main/Homeworks/Homework-3/crime-training-data")
test_df <- read.csv("https://raw.githubusercontent.com/waheeb123/Data-621/main/Homeworks/Homework-3/crime-evaluation-data")

```








