---
title: "exercises chapter 1 LMR"
author: "waheeb Algabri"
output:
  pdf_document: default
  html_document:
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(openintro)
```


**`1.The dataset teengamb concerns a study of teenage gambling in Britain. Make a numerical and graphical summary of the data, commenting on any features that you find interesting. Limit the output you present to a quantity that a busy reader would find sufficient to get a basic understanding of the data.`**

```{r}
library(faraway)
data(teengamb)
head(teengamb)
```

```{r}
summary(teengamb)
```

```{r}
# Graphical summary
par(mfrow=c(2, 3))
hist(teengamb$income, main="Income Distribution", xlab="Income")
hist(teengamb$verbal, main="Verbal Score Distribution", xlab="Verbal Score")
hist(teengamb$gamble, main="Gambling Behavior Distribution", xlab="Gambling Behavior")
boxplot(income ~ sex, data=teengamb, main="Income by Gender", xlab="Gender", ylab="Income")
boxplot(gamble ~ sex, data=teengamb, main="Gambling Behavior by Gender", xlab="Gender", ylab="Gambling Behavior")
```

2.The dataset uswages is drawn as a sample from the Current Population Survey in 1988. Make a numerical and graphical summary of the data as in the previous question

```{r}
data("uswages")
head(uswages)
```

```{r}
summary(uswages)
```

**2-The dataset uswages is drawn as a sample from the Current Population Survey in 1988. Make a numerical and graphical summary of the data as in the previous question.**

```{r}
head(uswages)
```

```{r}
summary(uswages)

```

```{r}
hist(uswages$wage, main="Wage Distribution", xlab="Wage")

```

```{r}
boxplot(uswages$wage, main="Boxplot of Wage", ylab="Wage")

```

```{r}
plot(uswages$exper, uswages$wage, main="Scatterplot of Wage vs Experience", xlab="Experience", ylab="Wage")

```

```{r}
barplot(table(uswages$educ), main="Barplot of Education Levels", xlab="Education Level", ylab="Frequency")

```

```{r}
ggplot(uswages, aes(x = exper, y = wage)) +
  geom_point() +
  labs(title = "Scatterplot of Wage vs Experience", x = "Experience", y = "Wage")

```

**3-The dataset 'prostate' is derived from a study involving 97 men diagnosed with prostate cancer, all of whom were scheduled to undergo a radical prostatectomy. Provide a numerical and graphical summary of the data, similar to the previous question.**

```{r}
head(prostate)
```

```{r}
summary(prostate)

```

Histogram for lcavol:

```{r}
hist(prostate$lcavol, main="lcavol Distribution", xlab="lcavol")

```

Boxplot for lcavol:

```{r}
boxplot(prostate$lcavol, main="Boxplot of lcavol", ylab="lcavol")

```

Scatterplot for lcavol vs. lpsa:

```{r}
plot(prostate$lcavol, prostate$lpsa, main="Scatterplot of lcavol vs lpsa", xlab="lcavol", ylab="lpsa")

```

```{r}
barplot(table(prostate$gleason), main="Barplot of Gleason Score", xlab="Gleason Score", ylab="Frequency")

```

```{r}

ggplot(prostate, aes(x = lcavol, y = lpsa)) +
  geom_point() +
  labs(title = "Scatterplot of lcavol vs lpsa", x = "lcavol", y = "lpsa")

```

**4-The dataset sat comes from a study entitled “Getting What You Pay For: The Debate Over Equity in Public School Expenditures.” Make a numerical and graphical summary of the data as in the first question.**

```{r}
head(sat)
```

```{r}
summary(sat)

```

```{r}
hist(sat$verbal, main="Verbal Score Distribution", xlab="Verbal Score")
hist(sat$math, main="Math Score Distribution", xlab="Math Score")
hist(sat$total, main="Total Score Distribution", xlab="Total Score")

```

```{r}
boxplot(sat$expend, main="Boxplot of Expenditure", ylab="Expenditure")

```

Scatterplot of Expenditure vs Salary

```{r}
plot(sat$expend, sat$salary, main="Scatterplot of Expenditure vs Salary", xlab="Expenditure", ylab="Salary")

```

```{r}
ggplot(sat, aes(x = expend, y = salary)) +
  geom_point() +
  labs(title = "Scatterplot of Expenditure vs Salary", x = "Expenditure", y = "Salary")

```

**5-The dataset divusa contains data on divorces in the United States from 1920 to 1996. Make a numerical and graphical summary of the data as in the first question.**

```{r}
head(divusa)
```
```{r}
summary(divusa)
```

```{r}
hist(divusa$divorce, main="Divorces Distribution", xlab="Number of Divorces")

```

```{r}
plot(divusa$year, divusa$divorce, type="l", main="Divorces Over Years", xlab="Year", ylab="Number of Divorces")

```

```{r}
plot(divusa$unemployed, divusa$femlab, main="Scatterplot of Unemployed vs Femlab", xlab="Unemployed", ylab="Femlab")

```


```{r}
library(ggplot2)
ggplot(divusa, aes(x = unemployed, y = femlab)) +
  geom_point() +
  labs(title = "Scatterplot of Unemployed vs Femlab", x = "Unemployed", y = "Femlab")

```



