---
title: "exercises chapter 1 LMR"
author: "Waheeb Algabri"
date: "2024-01-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


**1.The dataset teengamb concerns a study of teenage gambling in Britain. Make a numerical and graphical summary of the data, commenting on any features that you
 find interesting. Limit the output you present to a quantity that a busy reader would
find sufficient to get a basic understanding of the data.**


```{r}
library(faraway)
data(teengamb)
head(teengamb)
```

```{r}
summary(teengamb)
```

```{r}
# Graphical summary
par(mfrow=c(2, 3))
hist(teengamb$income, main="Income Distribution", xlab="Income")
hist(teengamb$verbal, main="Verbal Score Distribution", xlab="Verbal Score")
hist(teengamb$gamble, main="Gambling Behavior Distribution", xlab="Gambling Behavior")
boxplot(income ~ sex, data=teengamb, main="Income by Gender", xlab="Gender", ylab="Income")
boxplot(gamble ~ sex, data=teengamb, main="Gambling Behavior by Gender", xlab="Gender", ylab="Gambling Behavior")
```


2.The dataset uswages is drawn as a sample from the Current Population Survey in
1988. Make a numerical and graphical summary of the data as in the previous question

```{r}
data("uswages")
head(uswages)
```

```{r}
summary(uswages)
```















