title: "Homework-4"
author: "Waheeb Algabri, Joe Garcia, Lwin Shwe, Mikhail Broomes"
output:
  html_document:
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
  pdf_document: default
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)

library(tidyverse)
library(modelr)
library(DataExplorer)
library(correlationfunnel)
library(caret)
library(knitr)
library(confintr)
library(psych)
library(car)
library(corrplot)
library(RColorBrewer)
library(MASS)
select <- dplyr::select
library(glmtoolbox)
library(cowplot)
library(pROC)
library(bestglm)

```


## Multiple Linear Regression and Binary Logistic Regression 

In this homework assignment, you will explore, analyze and model a data set containing approximately 8000
records representing a customer at an auto insurance company. Each record has two response variables. The
first response variable, TARGET_FLAG, is a 1 or a 0. A “1” means that the person was in a car crash. A zero
means that the person was not in a car crash. The second response variable is TARGET_AMT. This value is zero
if the person did not crash their car. But if they did crash their car, this number will be a value greater than zero.
Your objective is to build multiple linear regression and binary logistic regression models on the training data
to predict the probability that a person will crash their car and also the amount of money it will cost if the person
does crash their car. You can only use the variables given to you (or variables that you derive from the variables
provided). Below is a short description of the variables of interest in the data set:


```{r}
train_df <-  read.csv("https://raw.githubusercontent.com/waheeb123/Data-621/main/Homeworks/Homework%204/insurance_training_data.csv")
test_df <- read.csv("https://raw.githubusercontent.com/waheeb123/Data-621/main/Homeworks/Homework%204/insurance-evaluation-data.csv")
```


```{r}
dim(train_df)
```
The dataset consists of 466 observations of 26 variables. There are 24 predictor variables and two response variable (`TARGET_FLAG` and `TARGET_AMT`).


```{r}
train_df |>
    glimpse()
```
There are two target variables `TARGET_FLAG` and `TARGET_AMT`. `TARGET_AMT` is a a continuous variable while `TARGET_FLAG` is a dummy variable which we will change to a factor to ensure R treats our variable as a categorical variable. 

```{r}
train_df <- train_df |>
    mutate(TARGET_FLAG = as.factor(TARGET_FLAG))
```


We will now take a look at summary statistics 

```{r}
summary(train_df)

```



